Synthesizing Symmetric Lenses

### 2018-07-09 papers/SymmetricLenses/main.tex
stuff

In this paper, we have designed and implemented a new, pragmatic
formulation of symmetric lenses as well as new program synthesis
techniques.  In this section, we consider the relationship with
classical symmetric lenses in depth and also comment on related
program synthesis techniques.

\subsection{Relationship with Classical Symmetric Lenses}
\label{sec:relationship}
%\paragraph*{Formalization}
%We have reformulated the problem addressed by classical
%symmetric lenses with simple symmetric lenses.
A classical symmetric lens
$\Lens$ between $X$ and $Y$ consists of 4 components: a complement $C$, a
designated element $init \in C$, and two functions,
$\PutRSym$ and $\PutLSym$, that propagate changes in one format to the other.
\begin{enumerate}
\item $\PutRSym \OfType X \times C \rightarrow Y \times C$
\item $\PutLSym \OfType Y \times C \rightarrow X \times C$
\end{enumerate}
In this formulation, data unique to each side are stored in the complement.
When one format is edited, the \PutR
or \PutL function stitches together the edited data with data
stored in the complement.  The  $init$ element is used as the
initial value of $C$ and specifies default behavior when data is missing.
For instance, to implement the
scenario in Figure~\ref{fig:minimized-representations}, the complement
would consist of a list of pairs of salary and company name.

@@

Two classical symmetric lenses are equivalent if they output the same
formats given any sequence of edits. Formally, given a lens
$\Lens \in X \leftrightarrow Y$,
an \emph{edit} for $\Lens$ is a member of $X + Y$. Consider
the function $apply$, which, given a lens and an element of that lens's
complement, is a function from sequences of edits to sequences of edits. If
$apply(\Lens,c,es) = es'$, then that means, starting with complement $c$, after
edit $es_i$, the lens $\Lens$ synchronizes the other data to $es'_i$.
\[
  \inferrule
  {
  }
  {
    apply(\Lens,c,[]) = []
  }
\]
\[
  \inferrule
  {
    \Lens.putr(x,c) = (y,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InLOf{x})::es) = (\InROf{y})::es'
  }
\]
\[
  \inferrule
  {
    \Lens.putl(y,c) = (x,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InROf{y})::es) = (\InLOf{x})::es'
  }
\]

Two lenses, $\Lens_1$ and $\Lens_2$, are equivalent if
$apply(\Lens_1,\Lens_1.init,es) = apply(\Lens_2,\Lens_2.init,es)$ for all edit
sequences $es$.

By describing our lenses in this complement style, we can express an incredibly
wide class\bcp{where is the evidence that this class of lenses is {\em
    usefully} wider than simple symmetric lenses?} of lenses, as these complements permit storing any amount of
information in them.  However, they pose problems when implementing them in
real-world systems and in synthesizing them.

\paragraph*{Issues with Classical Symmetric Lenses}
\iffalse
Using symmetric lenses for
synchronization forces us to maintain not just two files that we aim to
sycnrhonize, but three. \bcp{I don't find it unintuitive, and I think we
  should be careful not to oversell the difficulty of storing the
  complements.  It requires more engineering, but it's not rocket
  science.}This is unintuitive, and creates many problems in 
practice. For example, consider the situation where I am synchronizing a crontab
and job manager file. I am synchronizing across computers, so I must either
maintain a copy of the complement on each computer, and keep them synchronized,
or figure out a computer to hold the complement, and make sure all edits go
through that complement. Furthermore, if I want to change the synchronization to
a different computer, I don't need to merely make sure that my files are moved
to the computer, I also have to make sure that I correctly move the complement,
and hook up that complement to be used for that synchronization task. This is a
huge divergence from other synchronization utilities\bcp{No.  For example,
  Unison maintains a database on each replica recording, for each file, what
its hash was after the last time the two replicas were synchronized.}, where all
that needs to be 
done is maintain the files. Using those for synchronization, edits can be sent
over the wire, and the appropriate function can stitch together those edits, no
additional state needed.
\fi
For one, using classical symmetric lenses for synchronization forces us to maintain three files instead of two, which requires more engineering.
The more significant issues with classical symmetric lenses arise in synthesis.
Firstly, because each lens has a custom complement, we can no longer
specify the put functions though mere input/output examples.  Instead, we have
to specify the behavior through the observable behavior of edit sequences.
Furthermore, there are lenses that require arbitrarily long edit sequences to
differentiate between: consider the lens that keeps two formats equal for the
first $k$ non-identity edits, then keeps them completely unsynchronized for all
future edits.  This requires an edit sequence of $k$ edits to differentiate
between this lens and the identity lens.  Simple symmetric lenses avoid this
problem by not allowing such lenses: we can't keep track of the number of
non-identity edits performed unless there are explicit fields in the data
formats keeping track of them.  While we are reducing the expressivity of our
lenses with this restriction, we feel the benefits outweigh the costs.

\paragraph*{Relation with Simple Symmetric Lenses} To formalize how simple
symmetric lenses work as synchronization tools, we define an $apply$
function on them, much like we define an $apply$ on classical symmetric lenses. The
function $apply$, given a simple symmetric lens and an optional pair of elements of the
source and target, is a function from sequences of put objects to sequences of
put objects. If $apply(\Lens,\None,es) = es'$, then that means, starting with no
prior data, after edit $es_i$, the lens \Lens synchronizes the other data to
$es_i'$. If $apply(\Lens,\SomeOf{(x,y)},es) = es'$, then that means, starting with
data $x$ and $y$ on the left and right, respectively, after edit $es_i$, the
lens \Lens synchronizes the other data to $es_i'$.

@@

We've reformulated symmetric lenses in a style similar to existing asymmetric
lens formulations. However, some symmetric lenses certainly are inexpressible in
the simple symmetric lens formulation. We formalize which are expressible
with a restriction on symmetric lenses. We define \emph{forgetful symmetric
  lenses} to be symmetric lenses which satisfy the following additional laws:
\begin{equation}
  \tag{\ForgetfulRL}
  \begin{mathprooftree}
    \AxiomC{$\Lens.putr(x,c_1) = (\_,c_1')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putr(x,c_2) = (\_,c_2')$}
    \AxiomC{$\Lens.putl(y,c_1') = (\_,c_1'')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putl(y,c_2') = (\_,c_2'')$}
    \def\extraVskip{2pt}
    \singleLine
    \BinaryInfC{$c_1'' = c_2''$}
  \end{mathprooftree}
\end{equation}
\begin{equation}
  \tag{\ForgetfulLR}
  \begin{mathprooftree}
    \AxiomC{$\Lens.putl(y,c_1) = (\_,c_1')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putl(y,c_2) = (\_,c_2')$}
    \AxiomC{$putr(x,c_1') = (\_,c_1'')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putr(x,c_2') = (\_,c_2'')$}
    \def\extraVskip{2pt}
    \singleLine
    \BinaryInfC{$c_1'' = c_2''$}
  \end{mathprooftree}
\end{equation}

Intuitively, this says that the complements are uniquely determined by the most
recently input $x$ and $y$. This corresponds closely with simple symmetric
lenses, where all state is maintained by the current state of the $x$ and $y$
data. Forgetful symmetric lenses express exactly the same $apply$ function as
simple symmetric lenses.

@@

\paragraph*{Relationship with Asymmetric Lenses}
Formally, an {\em asymmetric lens} $ \ell : S \Leftrightarrow V$ is a triple of functions $\ell.\get : S \longrightarrow V$, $\ell.\pput : V \longrightarrow S \longrightarrow S$ and $\ell.\create : V \longrightarrow S$ satisfying the following laws \cite{Focal2005-long2}:
\begin{align*}
\ell.\get \; (\ell.\pput \; s \; v) &= v \tag{PUTGET}\\
\ell.\pput \; s \; (\ell.\get \; s) &= s \tag{GETPUT}\\
\ell.\get \; (\ell.\create \; v) &= v \tag{CREATEGET}
\end{align*}
Despite being less expressive than symmetric lenses, simple symmetric lenses are
strictly more expressive than classical asymmetric lenses.

@@

\subsection{Data Transformation Synthesis}
Over the past decade, the programming languages community has explored
the synthesis of programs from a wide variety of angles.  One of
the key ideas is typically to narrow the program search space by focusing
on a specific domain, as we have done here.
%, and as has been advocated
%more generally under the rubric of syntax-guided synthesis~\cite{sygus}.

>>>>

In this paper, we have designed and implemented a new, pragmatic
formulation of symmetric lenses as well as new program synthesis
techniques.  In this section, we analyze the relationship
between our simple symmetric
lenses and two previous lens languages:  classical symmetric lenses
and asymmetric lenses.  We also comment on related
program synthesis techniques.

\subsection{Relationship with Classical Lens Languages}
\label{sec:relationship}
%\paragraph*{Formalization}
%We have reformulated the problem addressed by classical
%symmetric lenses with simple symmetric lenses.

\paragraph*{Symmetric Lenses}
A classical symmetric lens
$\Lens$ between $X$ and $Y$ consists of 4 components: a complement $C$, a
designated element $init \in C$, and two functions,
$\PutRSym$ and $\PutLSym$, that propagate changes in one format to the other.
\begin{enumerate}
\item $\PutRSym \OfType X \times C \rightarrow Y \times C$
\item $\PutLSym \OfType Y \times C \rightarrow X \times C$
\end{enumerate}
In this formulation, data unique to each side are stored in the complement.
When one format is edited, the \PutR
or \PutL function stitches together the edited data with data
stored in the complement.  The  $init$ element is used as the
initial value of $C$ and specifies default behavior when data is missing.
For instance, to implement the
scenario in Figure~\ref{fig:minimized-representations}, the complement
would consist of a list of pairs of salary and company name.

@@

Two classical symmetric lenses are equivalent if they output the same
formats given any sequence of edits. Formally, given a lens
$\Lens \in X \leftrightarrow Y$,
an \emph{edit} for $\Lens$ is a member of $X + Y$. Consider
the function $apply$, which, given a lens and an element of that lens's
complement, is a function from sequences of edits to sequences of edits. If
$apply(\Lens,c,es) = es'$, then that means, starting with complement $c$, after
edit $es_i$, the lens $\Lens$ synchronizes the other data to $es'_i$.
\[
  \inferrule
  {
  }
  {
    apply(\Lens,c,[]) = []
  }
\]
\[
  \inferrule
  {
    \Lens.putr(x,c) = (y,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InLOf{x})::es) = (\InROf{y})::es'
  }
\]
\[
  \inferrule
  {
    \Lens.putl(y,c) = (x,c')\\
    apply(\Lens,c',es) = es'
  }
  {
    apply(\Lens,c,(\InROf{y})::es) = (\InLOf{x})::es'
  }
\]
\noindent
Two lenses, $\Lens_1$ and $\Lens_2$, are equivalent if
$apply(\Lens_1,\Lens_1.init,es) = apply(\Lens_2,\Lens_2.init,es)$ for all edit
sequences $es$.

%% By describing our lenses in this complement style, we can express an incredibly
%% wide class\bcp{where is the evidence that this class of lenses is {\em
%%     usefully} wider than simple symmetric lenses?} of lenses, as these complements permit storing any amount of
%% information in them.  However, they pose problems when implementing them in
%% real-world systems and in synthesizing them.

%\paragraph*{Issues with Classical Symmetric Lenses}
\iffalse
Using symmetric lenses for
synchronization forces us to maintain not just two files that we aim to
sycnrhonize, but three. \bcp{I don't find it unintuitive, and I think we
  should be careful not to oversell the difficulty of storing the
  complements.  It requires more engineering, but it's not rocket
  science.}This is unintuitive, and creates many problems in 
practice. For example, consider the situation where I am synchronizing a crontab
and job manager file. I am synchronizing across computers, so I must either
maintain a copy of the complement on each computer, and keep them synchronized,
or figure out a computer to hold the complement, and make sure all edits go
through that complement. Furthermore, if I want to change the synchronization to
a different computer, I don't need to merely make sure that my files are moved
to the computer, I also have to make sure that I correctly move the complement,
and hook up that complement to be used for that synchronization task. This is a
huge divergence from other synchronization utilities\bcp{No.  For example,
  Unison maintains a database on each replica recording, for each file, what
its hash was after the last time the two replicas were synchronized.}, where all
that needs to be 
done is maintain the files. Using those for synchronization, edits can be sent
over the wire, and the appropriate function can stitch together those edits, no
additional state needed.
\fi
While classical symmetric are more expressive than simple symmetric
lenses, it is unclear whether this additional expressiveness is useful in
practice.  Moreover, classic symmetric lenses come with 
some disadvantages.
For one, an implementation of classical symmetric lenses
would need to squirrel away the complement in
an auxiliary file somwhere, which likely requires more engineering.
However, more significant issues arise in synthesis.
First, because each lens has a custom complement, one can no longer
specify the put functions through input/output examples alone.  One
alternative would be to enrich specifications with edit sequentials;
another would be to specify the structure of complements explicitly
(though the latter would be somewhat akin to specifying the internal
state of a program).  In either case, the complexity of the specifications
increases.
%% addition, users would have to specify some auxilary 
%% to specify the behavior through the observable behavior of edit sequences.
%% Second, there are lenses that require arbitrarily long edit sequences to
%% differentiate between: consider the lens that keeps two formats equal for the
%% first $k$ non-identity edits, then keeps them completely unsynchronized for all
%% future edits.  This requires an edit sequence of $k$ edits to differentiate
%% between this lens and the identity lens.  Simple symmetric lenses avoid this
%% problem by disallowing such lenses: we can't keep track of the number of
%% non-identity edits performed unless there are explicit fields in the data
%% formats keeping track of them.  While we are reducing the expressivity of our
%% lenses with this restriction, we feel the benefits outweigh the costs.

\paragraph*{Relation with Simple Symmetric Lenses} To formalize how simple
symmetric lenses work as synchronization tools, we define an $apply$
function on them, much like one defines an $apply$ on classical symmetric lenses:
Given a simple symmetric lens and an optional pair of 
source and target strings,
$apply$ is a function from sequences of put objects to sequences of
put objects. If $apply(\Lens,\None,es) = es'$, then starting with no
prior data, after edit $es_i$, the lens \Lens synchronizes the other data\dpw{Is the ``other data'' $es$? or is it something else? If it is $es$ then we can say ($es$) in the sentence.} to
$es_i'$. \dpw{I wasn't sure what a ``put object'' was.  Is it a function? (If so, perhaps
call it a function and give its type?)} If $apply(\Lens,\SomeOf{(x,y)},es) = es'$, then that means, starting with
data $x$ and $y$ on the left and right, respectively, after edit $es_i$, the
lens \Lens synchronizes the other data to $es_i'$.

@@

We have reformulated symmetric lenses in a style similar to existing asymmetric
lens formulations. However, some symmetric lenses certainly are inexpressible in
the simple symmetric lens formulation. We formalize which are expressible
with a restriction on symmetric lenses. We define \emph{forgetful symmetric
  lenses} to be symmetric lenses that satisfy the following additional laws.
\begin{equation}
  \tag{\ForgetfulRL}
  \begin{mathprooftree}
    \AxiomC{$\Lens.putr(x,c_1) = (\_,c_1')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putr(x,c_2) = (\_,c_2')$}
    \AxiomC{$\Lens.putl(y,c_1') = (\_,c_1'')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putl(y,c_2') = (\_,c_2'')$}
    \def\extraVskip{2pt}
    \singleLine
    \BinaryInfC{$c_1'' = c_2''$}
  \end{mathprooftree}
\end{equation}
\begin{equation}
  \tag{\ForgetfulLR}
  \begin{mathprooftree}
    \AxiomC{$\Lens.putl(y,c_1) = (\_,c_1')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putl(y,c_2) = (\_,c_2')$}
    \AxiomC{$putr(x,c_1') = (\_,c_1'')$}
    \def\extraVskip{.5pt}
    \noLine 
    \UnaryInfC{$\Lens.putr(x,c_2') = (\_,c_2'')$}
    \def\extraVskip{2pt}
    \singleLine
    \BinaryInfC{$c_1'' = c_2''$}
  \end{mathprooftree}
\end{equation}

Intuitively, these equations state that complements are uniquely determined by the most
recently input $x$ and $y$. Hence, such lenses correspond closely \dpw{replace ``closely'' with
``exactly''?} with simple symmetric
lenses, where all state is maintained by the current state of the $x$ and $y$
data. Forgetful symmetric lenses express exactly the same $apply$ function as
simple symmetric lenses.

@@

\paragraph*{Asymmetric Lenses}
Formally, an {\em asymmetric lens} $ \ell : S \Leftrightarrow V$ is a triple of functions $\ell.\get : S \longrightarrow V$, $\ell.\pput : V \longrightarrow S \longrightarrow S$ and $\ell.\create : V \longrightarrow S$ satisfying the following laws \cite{Focal2005-long2}:
\begin{align*}
\ell.\get \; (\ell.\pput \; s \; v) &= v \tag{PUTGET}\\
\ell.\pput \; s \; (\ell.\get \; s) &= s \tag{GETPUT}\\
\ell.\get \; (\ell.\create \; v) &= v \tag{CREATEGET}
\end{align*}
Despite being less expressive than symmetric lenses, simple symmetric lenses are
strictly more expressive than classical asymmetric lenses.

@@

\subsection{Data Transformation Synthesis}
Over the past decade, the programming languages community has explored
the synthesis of programs from a wide variety of angles.  One of
the key ideas is typically to narrow the program search space by focusing
on a specific domain, and imposing constraints on syntax~\cite{sygus} or
typing~\cite{augustsson-2004,gvero-pldi-2013,osera+:pldi15,feser-pldi-2015,scherer-icfp-2015,frankle+:popl16,}, or both.
%, and as has been advocated
%more generally under the rubric of syntax-guided synthesis~\cite{sygus}.
